COLLISION RATES OF PLANETESIMALS NEAR MEAN-MOTION RESONANCES

### 2020-04-13 main.tex
Finish results section

We begin by making the assumption that a typical encounter between planetesimals in a disk is well-described by two-body 
dynamics. So long as the random velocities of planetesimals $v_{p}$ are large compared to the Hill velocity of a planetesimal $v_{h} 
= \sqrt{G m_{p} / r_{h}}$, where $r_{h}$ is the Hill radius, the gravitational field of the central star will play an insignificant role and 
can be safely ignored. In this case and assuming equal mass bodies, the collision rate is given by \citep{1967SvA....10..650S}

@@

The net result of this is that isolated MMRs will produce 'spikes' in the semi-major axis vs eccentricity distribution of a planetesimal 
disk. Following equation \ref{eq:tiss}, these spikes will tend to slightly curve away from the perturbing body. This tends to lower the surface 
density of the disk near the locations of resonances, while producing a pileup of bodies on eccentric orbits adjacent to the MMR. The 
enhanced surface density will increase the collision rate, while the larger eccentricites will lower it due to the decreased effectiveness of 
gravitational focusing. In addition, the MMRs will drive fast precession of the orbiting planetesimals, which effectively randomizes their 
longitudes of perihelion. In the case of a secularly forced disk, there will be an interface between aligned and randomly oriented orbits at 
the edge of the resonance. Due to the complicated dynamics that are likely at play near the edge of the resonant region, we turn to a direct 
N-body treatment to better understand how the collision rate varies in the vicinity of a MMR.

@@

@@

@@

The MMRs are most clearly visible when viewing the semimajor axis and eccentricity distribution of the planetesimals. This is shown 
in figure \ref{fig:ae}. The top panel corresponds to the CJ simulation and the bottom panel corresponds to EJ. The solid vertical lines 
in this plot represent the nominal locations of the resonances. Near the nominal resonance locations, the a-e distribution of the 
planetesimals exhibits well-defined spikes which curve slightly inward due to conservation of the Jacobi energy. In the EJ simulation, 
the eccentricity distribution of bodies near resonance extends below the initial forced value. This is a reflection of the fact that these 
bodies are undergoing libration, which drives periodic oscillations in both semimajor axis and eccentricity. The coupling between 
these two quantities is described by equation \ref{eq:tiss}. In the EJ simulation, the eccentricity remains enhanced between the 
MMRs due to the secular forcing by Jupiter. Because the strength of the resonances scales with the initial eccentricity of the planetesimals, 
the eccentricity spikes are larger in the EJ simulation. Additionally, the relative size of the 3:1 and 5:3 MMR spikes also grows because the 
strength of a MMR scales as $e^{k}$, where $k$ is the order of the resonance (citation needed).

The MMRs have one more effect, which is to drive precession of the longitudes of perihelion of planetesimals near 
resonance. This effect becomes important in the EJ simulation, in which the orbits of bodies are initially aligned with that of Jupiter 
(see section \ref{sec:sec_force}). As is shown in figure \ref{fig:long_ph}, the strong precession induced by the MMRs overcomes the 
secular forcing and randomizes the orbital orientations of bodies near the resonances. The dashed horizontal lines in figure 
\ref{fig:long_ph} indicate the libration width of a planetesimal with an eccentricity of $e_{forced}$ at each resonance, which 
approximately shows the zone of influence of each MMR. It is important to note that the collision rate predicted by equation
\ref{eq:safronov} assumes that the orbits of bodies are randomly oriented. Near the interface of the resonances, the N-body collision rate, 
which we will examine next, is likely not well-described by the Safronov rate.

@@

As we have already discussed, the Safronov collision rate given by equation \ref{eq:safronov} suffers from two important limitations. The
first is that the orbits of colliding bodies exhibit no preferred orientation. Second, as we will show below, the velocities of planetesimals 
near the MMRs follows a bimodal distribution. This makes it difficult to characterize the planetesimals with a single 'typical' encounter 
velocity and therefore predicting the collision rate using a statistical approach is far from straightforward.

Figure \ref{fig:coll_hist_a} shows the semimajor axis of collisions resolved in the CJ (top) and EJ simulations (bottom). We define 
'semimajor axis of a collision' as the semimajor axis of the first of the two planetesimals participating in a collision at the 
moment that it happens. To ensure that the definition of the first and second collider chosen by {\sc ChaNGa} does not influence 
these results, we also tried constructing figure \ref{fig:coll_hist_a} using the second collider and found no qualitative differences. 
Although both simulations were integrated and collisions were resolved for 5,000 years, we only show collisions that occurred during 
the last 3,000 years of the simulations. Before this time, the resonant features in the semimajor axis-eccentricity distribution (see 
figure \ref{fig:ae}) were still developing and had not yet reached a steady state (the libration timescale of the MMRs we are interested in turns out to be about 2,000 years). This results in transient features appearing in the collision profiles early on, which we exclude from figure \ref{fig:coll_hist_a} and all subsequent figures.

We find that the features present in figure \ref{fig:coll_hist_a} are highly sensitive to the choice of binning when constructing a standard histogram and instead model the occurrence of collisions as a function of semimajor axis using a kernel density estimate (KDE). We use the {\sc neighbors.KernelDensity} function from the {\sc sklearn} \citep{scikit-learn} package to construct our KDEs. For our kernel, we use a tophat function with a bandwidth of 0.01 AU.

Most notably, the MMRs appear to have a measurable effect on the local collision rates. This is in contrast to \citet{2000Icar..143...45R}, 
who simulated a similar setup and found no discernible features near the MMRs. We attribute the difference to a more careful treatment of 
timestepping in our simulations. Slightly interior to the 2:1 resonance, the collision rate increases by a factor of roughly 2. In the EJ 
simulation, there is a significant deficiency in the number of collisions near the 3:1 MMR. We attribute this to the fact that the strength of a 
second order MMR scales as $e^{2}$, while a first order MMR scales linearly with $e$. One would therefore expect the 3:1 resonance to 
have more of an influence on the collision rate in a secularly perturbed disk. Although viewing the collision rate as a function of semimajor 
axis demonstrates a clear connection between the resonances and the collision rate, this does not immediately indicate where the dust 
generated by collisions will end up in the disk.

@@

Operating under this assumption, a map of the relative concentration of second-generation dust can be constructed from the cylindrical distance at which the collisions occur. This is shown in the top panels of figure \ref{fig:coll_hist_r}, where the left side corresponds to the CJ simulation and the right side corresponds to EJ. As in figure \ref{fig:coll_hist_a}, a kernel density estimate of the number of collisions is constructed using a tophat kernel with a bandwidth of 0.01 AU. Features again appear near the 2:1 and 3:1 MMRs in both simulations, although the shape of the features are qualitatively different from figure \ref{fig:coll_hist_a}. Near the 2:1 resonance, the radial collision rate is noticably suppressed. This is mainly due to the reduced surface density of planetesimals near the center of the resonance, shown in the second row of figure \ref{fig:coll_hist_r}. The surface density in the EJ simulation near this resonance is less suppressed than the CJ simulation due to the enhanced eccentricity of planetesimals in the vicinity. This also has the effect of washing out the feature created in the collision rate profile due to the 2:1 MMR.

Additionally, a sharp, well-defined feature appears just exterior to the 3:1 MMR in the EJ simulation. A corresponding dip in the surface density at this same location suggests that the variation in the collision rate here is due to a lack of planetesimals in this location. Upon closer inspection, we find that the center of this feature lies at a a radial distance of $(1+e_{forced})a_{3:1}$, where $a_{3:1}$ is the nominal location of the 3:1 MMR and $e_{forced}$ is the forced eccentricity of a planetesimal at $a_{3:1}$. This distance corresponds to the aphelion distance of a planetesimal with these orbital parameters. Physically, the drop in the surface density here can be explained as follows: bodies near the 3:1 MMR experience strong precession, which randomizes their orbits and undergo libration which produces a large spread in their relative velocities. For a collection of randomly oriented eccentric orbits, there is less geometric overlap at aphelion, which lowers the surface density at this distance. Given the clear connection between the offset of this feature and the local forced eccentricity, we will examine the feasibility of using this feature to measure the eccentricity of the perturbing planet in section \ref{sec:dust} after we construct a dust emission map from the collision rate profile.

The bottom row of figure \ref{fig:coll_hist_r} demonstrates the difficulty with calculating the collision rate from the position and velocity distribution of particles. Typically, the rms velocity is defined as $\sqrt(\left<e^2\right> + \left<i^2\right>)$. In this case, the wings extending off of the resonances in the distance-eccentricity distribution makes the eccentricity dispersion distinctly bimodal and not well defined at nearly all radial distances. A naive calculation of the Safronov rate here would predict a collision rate near the resonances that is smaller than the actual value.

@@

%Next, we move on to examine the statistics of the collisions that occur. Figure \ref{fig:coll_hist_a} shows the distribution of semimajor axes of every collision that is resolved. Here, the semimajor axis of a collision is defined as the semimajor axis of the first body participating in the collision at the moment that it occurs. During a collision, the determination of which body is first and which is second is determined by the collision search tree in {\sc ChaNGa}. To ensure that this does not introduce any bias, we also constructed figure \ref{fig:coll_hist_a} using the orbital parameters of the second body and found no qualitative difference. By examining the semimajor axis distribution of the collisions, it is easy to see how the MMRs affect the collision rates. The dashed lines again represent the libration width of each resonance and are meant to show its range of influence. In both the circular and eccentric Jupiter simulations, the collision rate is clearly enhanced interior to the 2:1 MMR. The 3:1 MMR only appears to have a noticeable influence on the collision rate in the EJ simulation. In both cases, although the 5:3 MMR has a clear influence on the on the semimajor-axis eccentricity distribution, it does not leave a measureable imprint on the collision rate. It is important to point out, however, that the libration width of sufficiently eccentric particles near the 5:3 MMR may extend beyond the outer edge of the disk and introduce boundary effects.

%Although the semimajor axis of a particle at the moment of collision provides useful information about how the mean motion resonances affect the collision rate in the disk, this information alone is not enough to predict where the second generation dust that is produced will end up. To do so, we will operate under the assumption that dust grains produced by collision events will quickly couple to the gas and attain a circular orbit. This assumption, which will be further justified in section \ref{sec:dust}, allows to map the radial variation in the dust surface density to the radial position of the collisions. This is shown in figure \ref{fig:coll_hist_r}, where the histograms indicate the x and y coordinates of a body at the moment that it undergoes a collision.

%Explanation of offset of feature near 3:1 in EJ: orbital orientations randomized near MMR, most bodies have ecc of e_forced, even in resonance. Geometrically, less overlap at apo, so surface density drops and so does coll rate.
Most notably, the prominent features visible in figure \ref{fig:coll_hist_a} adjacent to the 2:1 MMR no longer appear. This is due to the 
fact that the orbits of bodies in the 2:1 MMR have a wide range of eccentricities and therefore have the potential to undergo 
collisions at a wide range of heliocentric distances. In the CJ simulation, there is a prominent dip in the collision rate near the 2:1 
MMR. This feature is still moderately visible in the EJ simulation, but the dip is much wider and shallower. The random orientations of 
the orbits inside of the resonance (see figure \ref{fig:long_ph}), combined with the forced eccentricites of planetesimals adjacent to 
the resonance weakens the dip near the 2:1 MMR seen in the left hand panel of figure \ref{fig:coll_hist_a}. Additionally, there is a 
double-pronged feature in the collision rate near the 3:1 MMR in the EJ simulation. The radial distance of these two prongs 
corresponds to the maximum radial excursion that a planetesimal at the 3:1 MMR with an eccentricity set by secular forcing with 
Jupiter undergoes. (Need more explanation of why this happens)

As is seen in figure \ref{fig:long_ph}, the relatively fast precession rates induced by the mean motion resonances overpower the 
secular forcing by Jupiter and randomize the orientation of orbits near the nominal resonance locations. Of course, this difference in 
longitude of perihelion of bodies inside and outside of resonance is only present in the EJ simulation because the Jupiter imposes 
zero forced eccentricity on the disk when it has a circular orbit. Figure \ref{fig:m_hist} provides another view of the difference in 
collision statistics inside and outside and of resonance. This shows the mean anomalies of bodies at the moment that they undergo a 
collision. The top row corresponds to the EJ simulation and the bottom rom corresponds to the CJ simulation. The left panels 
correspond to collisions that occur inside of the 2:1 MMR and the right panels are for collisions that occur with a semimajor axis 
between 2.7 and 2.8 AU, which is meant to well away from the influence of any resonances. The semimajor axis window size in the 
right hand panel is chosen so that both histograms contain the same total number of collisions.

The qualitative differences between these four panels can be explained in the following way: for $\approx$ 100 km bodies, 
gravitational focusing plays a significant role in setting the collision rate. In regions of the disk where the rms eccentricity is small 
(which corresponds to a small velocity dispersion), the collision rate should be enhanced (see equation \ref{eq:safronov}). Inside of 
resonances, planetesimals undergo large amplitude libration which pumps up the eccentricity dispersion. This effect is visible in 
figure \ref{fig:ae} for both the EJ and CJ simulation. This implies that the collision rate should be geometric, and not depend on the 
velocity dispersion near resonances. When this is true, collisions should preferentially occur near perihelion. The reason for this is 
that in the geometric limit, the collision probability scales with the ratio between the geometric cross section of the two planetesimals 
and the surface area of a sphere with radius $r$, where $r$ is the distance between the bodies and the central star. This ratio, and 
therefore the collision probability, is largest when $r$ is minimized (see \citet{2003AJ....125.2692L}). The fact both of the left-hand 
panels in figure \ref{fig:m_hist} show a distinct lack of collisions near aphelion is consistent with this explanation.

Note that this trend is not present in the right-hand panels of figure \ref{fig:m_hist}. In the case with Jupiter on a circular orbit (bottom right panel), there is a spike in the collision rate both at perihelion and aphelion. The perihelion spike is still due to the geometric argument presented in the previous paragraph. The aphelion spike is due to the effect of gravitational focusing. At aphelion, bodies achieve their minimum speed. This lowers the relative velocity between planetesimals, which according to equation \ref{eq:safronov} should enhance the collision rate. Note that the collision rate does not exhibit much dependence on mean anomaly outside of resonance in the case with Jupiter on an eccentric orbit (top right panel). This is due to the secular forcing by Jupiter, which aligns the longitude of perihelia of adjacent planetesimals. The result of this is that bodies maintain similar relative speeds over the course of an orbit, even though the orbit is eccentric. This lack of variation in the relative speeds makes collisions equally likely at any orbital phase.

Motivate the assumption that dust produced in collisions immediately goes onto circular orbit

Discuss figure 5, hilight differences between surface density distribution and collision histogram. Calculation of safronov rate is hard because velocity distribution is bimodal in many places. Key features to point out: 2:1 MMR shows up well in CJ. Much more washed out in EJ, while 3:1 becomes more prominent. Also both features in EJ are shifted to the right of resonance location by a factor of (1+$e_{forced}$. This could be a viable way to measure ecc of the perturber.

\section{Observability of Collisions} \label{sec:dust}

Have aaron make a CASA dust emission map for the two histograms in figure 5. Fit a gaussian to the radial features and see if we can back out the eccentricity of the perturber and its mass. Also make 2d histogram showing how the perturbers eccentricity (which sets the forced eccentricity) and mass sets the libration width.

\section{Measuring Properties of the Perturber} \label{sec:fitting}

>>>>

We begin by making the assumption that a typical encounter between planetesimals in a disk is well-described by two-body 
dynamics. So long as the random velocities of planetesimals $v_{p}$ are large compared to the Hill velocity of a planetesimal $v_{h} 
= \sqrt{G m_{p} / r_{h}}$, where $r_{h}$ is the Hill radius, the gravitational field of the central star will play an insignificant role during a 
encounter and can be safely ignored. In this case and assuming equal mass bodies, the collision rate is given by 
\citep{1967SvA....10..650S}

@@

The net result of this is that isolated MMRs will produce 'spikes' in the semi-major axis vs eccentricity distribution of a planetesimal 
disk. Equation \ref{eq:tiss} predicts that these spikes will tend to slightly curve away from the perturbing body. This lowers the surface 
density of the disk near the locations of resonances, while producing a pileup of bodies on eccentric orbits adjacent to the MMR. The 
enhanced surface density will increase the collision rate, while the larger eccentricites will lower it due to the decreased effectiveness of 
gravitational focusing. In addition, the MMRs cause the orbits of planetesimals to precess on relatively short timescales, which effectively 
randomizes their longitudes of perihelion. In a secularly forced disk, there will be an interface between aligned and randomly oriented orbits 
at the edge of the resonance. Due to the complicated dynamics that are at play in this region, we turn to a direct N-body treatment to better 
understand how the collision rate varies in the vicinity of a MMR.

@@

The MMRs are most clearly visible when viewing the semimajor axis and eccentricity distribution of the planetesimals. This is shown 
in figure \ref{fig:ae}. The top panel corresponds to the CJ simulation and the bottom panel corresponds to EJ. The solid vertical lines 
in this plot represent the nominal locations of the resonances. Near the nominal resonance locations, the a-e distribution of the 
planetesimals exhibits well-defined spikes which curve slightly inward due to conservation of the Jacobi energy. In the EJ simulation, 
the eccentricity distribution of bodies near resonance extends below the initial forced value. This is a reflection of the fact that the resonant 
angles of these bodies are undergoing libration, which drives periodic oscillations in both semimajor axis and eccentricity. The coupling 
between these two quantities is described by equation \ref{eq:tiss}. Because the strength of the resonances scales with the initial 
eccentricity of the planetesimals, the eccentricity spikes are larger in the EJ simulation. Additionally, the relative size of the 3:1 and 5:3 
MMR spikes also grows because the strength of a MMR scales as $e^{q}$, where $q$ is the order of the resonance 
\citep{1994PhyD...77..289M}.

@@

The MMRs have an additional important effect, which is to drive precession of the longitudes of perihelion of planetesimals. 
This effect becomes important in the EJ simulation because the orbits of bodies are initially aligned with that of Jupiter 
(see section \ref{sec:sec_force}). As is shown in figure \ref{fig:long_ph}, the strong precession induced by the MMRs overcomes the 
secular forcing and randomizes the orbital orientations of bodies near the resonances. The dashed horizontal lines in figure 
\ref{fig:long_ph} indicate the libration width (should probably show the equation I used to calculate this) of a planetesimal with an 
eccentricity of $e_{forced}$ at each resonance, which approximately shows the zone of influence of each MMR.

@@

@@

A prediction of the collision rate using the Safronov equation (eq \ref{eq:safronov}) and the positions and velocities of the planetesimals
suffers from an important limitation. As we will show below, the velocities of planetesimals near the MMRs does not follow a unimodal 
distribution. This makes it difficult to characterize the motion of the planetesimals with a single encounter velocity $v_{p}$ and so predicting 
the collision rate using a statistical approach is far from straightforward. This operation becomes even more complicated with a secularly 
perturbed disk where there is a transition between aligned and randomly oriented orbits near the interface of the resonances (see figure 
\ref{fig:long_ph}.

Figure \ref{fig:coll_hist_a} shows the semimajor axis of collisions resolved in the CJ (top) and EJ simulations (bottom). We define 
the semimajor axis of a collision as the semimajor axis of the first of the two planetesimals participating in a collision at the 
moment that it happens. To ensure that the definition of the first and second collider chosen by {\sc ChaNGa} does not influence 
these results, we also tried constructing figure \ref{fig:coll_hist_a} using the properties of the second collider and found no qualitative 
differences. Although both simulations were integrated and collisions were resolved for 5,000 years, we only show collisions that occurred 
during the last 3,000 years of the simulations. Before this time, the resonant features in the semimajor axis-eccentricity distribution (see 
figure \ref{fig:ae}) were still developing and had not yet reached a steady state (the libration timescale of the MMRs we are interested in 
turns out to be about 2,000 years). This results in transient features appearing in the collision profiles early on, which we exclude from 
figure \ref{fig:coll_hist_a} and all subsequent figures.

We find that the features present in figure \ref{fig:coll_hist_a} are highly sensitive to the choice of binning when constructing a standard 
histogram and instead model the occurrence of collisions as a function of semimajor axis using a kernel density estimate (KDE). We use 
the {\sc neighbors.KernelDensity} function from the {\sc sklearn} \citep{scikit-learn} package to construct our KDEs. For the kernel, we use 
a tophat function with a bandwidth of 0.01 AU.

Near some of the MMRs, there are noticeable suppressions or enhancements of the local collision rate. This contrasts with the findings of 
\citet{2000Icar..143...45R}, who simulated a similar setup and found no discernible features near the MMRs. We attribute the differences to 
a more careful treatment of timestepping in our simulations. Just interior to the 2:1 resonance, the collision rate is strongly enhanced. In the 
EJ simulation, there is a significant deficiency in the number of collisions near the 3:1 MMR. This feature, although still present, is much 
weaker in the CJ simulation. We attribute this to the fact that the strength of a second-order MMR scales as $e^{2}$, while a first order 
MMR scales with $e$ \citep{1994PhyD...77..289M}. One would therefore expect the 3:1 resonance to have more of an influence on the 
collision rate in a more eccentric disk. Although viewing the collision rate as a function of semimajor axis demonstrates a clear connection 
between the resonances and the collision rate, this does not immediately indicate where the dust generated by collisions will end up in the 
disk. (Why is the feature around the 3:1 MMR so different than the 2:1?)

\subsection{Where Does the Dust End Up?}

@@

Operating under this assumption, a map of the relative concentration of second-generation dust can be constructed from the cylindrical 
distance at which the collisions occur. This is shown in the top panels of figure \ref{fig:coll_hist_r}, where the left side corresponds to the CJ 
simulation and the right side corresponds to EJ. As in figure \ref{fig:coll_hist_a}, a KDE of the collision rate is constructed using a tophat 
kernel with a bandwidth of 0.01 AU.

In contrast to what was seen in figure \ref{fig:coll_hist_a}, the radial collision rate is now suppressed near the 2:1 MMR. The spike in the 
collision rate at semimajor axes interior to this location becomes washed out when viewed instead by cylindrical distance because the 
spread in the eccentricities of the colliding planetesimals is so large. Although many of the collisions happen at the same semimajor axis, 
they occur at very different distances from the central star. The second row of figure \ref{fig:coll_hist_r} shows the azimuthally averaged 
surface density of the planetesimal disk at the end of the simulations. These profiles were calculated using the {\sc PYNBODY} analysis 
package \citep{2013ascl.soft05002P}.

In both the CJ and EJ simulations, the local minima in the surface density profiles match well with the minima in the collision rates. 
Additionally, the minima in the EJ simulation are offset from the centers of the resonances. This can be explained as follows: for 
planetesimals affected by the MMRs, the typical encounter velocity is large and gravitational focusing is suppressed. Here, the collision 
cross section is geometric. In this limit, the collision probability scales with the ratio between the geometric cross section of the 
planetesimals and the surface area of a sphere with radius $r$, where $r$ is the distance between the bodies and the central star. This 
ratio, and therefore the collision probability, is largest when $r$ is minimized (see \citet{2003AJ....125.2692L}). Therefore, the collision 
probability for planetesimals affected by resonance is smallest at aphelion. This is also consistent with the reduced surface density exterior 
to the resonances.

Finally, the bottom row of figure \ref{fig:coll_hist_r} demonstrates the difficulty with calculating the collision rate from the position and velocity 
distribution of particles. At a given distance from the central star, the distribution of velocities (using eccentricity as a proxy for typical encounter 
velocity), is not unimodal. In addition, the wings extending off of the resonances in these plots show that planetesimals can interact and collide 
with bodies far from the centers of the resonances. The bump seen in the collision rate in the EJ simulation around 2.7 AU is like a result of the 
complicated interaction between planetesimals in the 3:1, 5:2 and 2:1 resonances.

So far, we have shown that mean-motion resonances with a perturbing giant planet will produce measureable radial variations in the collision rates across a disk of planetesimals, which should give rise to variations in the dust profile. In addition, distinctly different features are produced if the giant planet is on an eccentric, rather than a circular orbit. In the next section, we would like to determine whether these features can be used to place any constraints on the mass and eccentricity of the planet.

@@

\section{Influence of Jupiter's Mass and Eccentricity on the Radial Dust Profile}

\section{Observability of Dust} \label{sec:dust}

Aaron's CASA images go here
